Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-03 draft.tex
Moved cite commands

Klemm et al. \cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA).

@@

Steenwijk et al. \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.

@@

Turkay et al. \cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.

@@

Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.

@@

Dai and Gahegan \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.

@@

Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.

@@

Niemann et al. \cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.

@@

Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data.
%%
Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, get too complex, analyzing these decision rules becomes complex.
%%
Sekhavat and Hoeber \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

@@

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

As described in a prior work \cite{Klemm2014VIS}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.

>>>>

Klemm et al. propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA) \cite{Klemm2014VIS}.

@@

Steenwijk et al. propose a coordinated linked view system for both image-related and non-image data \cite{Steenwijk}.

@@

Turkay et al. follow a similar approach by deriving descriptive data metrics from image data \cite{Turkay}.

@@

Zhang et al. analyze subject groups in a web-based linked view system \cite{Zhang}.

@@

Dai and Gahegan defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines \cite{Dai}.

@@

Gla{\ss}er et al. use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue \cite{Glasser2013}.

@@

Niemann et al. assess hepatic steatosis (fatty liver disease) risk factors using decision trees \cite{Niemann2014}.

@@

Pinhero et al. applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
%%
Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, get too complex, analyzing these decision rules becomes complex.
%%
Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

@@

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
